---
title: "Defining and using objects of class SURVIVAL"
author: John Aponte
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Defining and using objects of class SURVIVAL}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Here we present examples on how to construct and use objects of the
class SURVIVAL.

The function `s_factory(s_family,...)` is a function that call the
constructor of the family of distribution. Each family has it own set of
parameters. As the factories implement polymorphic behaviour according
to the parameters given, it is not possible to partial match the name of
the parameters and they need to be spell correctly. If an error on
processing the parameters, the factory return a message with the set of
parameters accepted for that factory.

Once an object of a class SURVIVAL is instantiated, it has access to the
following set of methods:

-   `sfx(SURVIVAL, t)` for the survival (proportion of the population
    free of events) at time `t`
-   `hfx(SURVIVAL, t)` for the hazard at time `t`
-   `Cum_Hfx(SURVIVAL, t)` for the cumulative hazard at time `t`
-   `InvCum_Hfx(SURVIVAL, H)` for inverse of the cumulative hazard `H`
-   `rsurv(SURVIVAL, n)` for the generation of `n` random survival times
    from the distribution
-   `rsurvhr(SURVIVAL, hr)` for the generation of random survival times
    with a hazard ratio `hr`

Instead of using the helper functions to call this methods, the methods
can be called directly from the object as:

-   `obj <- s_factory(s_family, ...)` or `obj <- s_family(...)`

-   `obj$sfx(t)`

-   `obj$hfx(t)`

-   `obj$Cum_Hfx(t)`

-   `obj$InvCum_Hfx(H)`

-   `obj$rsurv(n)`

-   `obj$rsurvhr(hr)`

In addition, the following functions help to plot the distributions

-   `plot(SURVIVAL)` a generic S3 method that calls the
    `plot_survival()` function

-   `plot_survival(SURVIVAL, timeto, main)` which plots the survival,
    hazard, cumulative hazard and the inverse cumulative hazard
    functions from 0 to `timeto`. An optional title can be specified
    with the `main` parameter

-   `ggplot_survival_random(SURVIVAL, timeto, subjects, nsim, alpha = 0.1)`
    produces a ggplot of Kaplan-Meier curve and Cumulative Hazard for
    `nsim` simulations for the a study with `subjects` number of
    subjects. The optional paramerter `alpha` defines the transparency
    of each simulation in the graph. In addition of the simulations, the
    graph also present the calculated survival and cumulative hazard
    function of the distribution, to evaluate how good the simulations
    are compared with the real values.
    
-   `plot_compare(SURVIVAL1, SURVIVAL2, timeto)` produce a comparison of the
     functions of two SURVIVAL objects.


The simulation of survival times and survival times with hazard
    ratios follow the method describe by @bender2003

```{r setup}
library(survobj)
library(survival)
library(ggplot2)
```

## Exponential Distribution

The canonical parameter of the exponential distribution is called
`lambda` and represents a constant hazard over time. The units of
`lambda` define the units of time for a distribution. For example if
`lambda = 3` is used to represent the probability of having 3 events in
1 year, the survival function `sfx(SURVIVAL, 1)` calculate the
proportion of the population free of events at 1 year.

The distribution can be defined also with the proportion of the
population free of events (`surv`) at time `t` or the proportion of the
population with events (`fail`) at time `t`

```{r exponential, fig.height=6, fig.width=7, fig.align='center'}
# Instanciate an object of class SURVIVAL with the Exponential distribution
obj1 <- s_factory(s_exponential, lambda = 3)
obj1

# Survival at time 1
sfx(obj1,1)

# Hazard at time 1
hfx(obj1,1)

# Cumulative hazard at time 1
Cum_Hfx(obj1,1)

# Inverse of the cumulative hazard 0.6
invCum_Hfx(obj1, 0.6)

# Plot of the distribution
plot(obj1)
```

The next set of examples show how to define an exponential distribution
based on the surviving or failing proportion at time `t`

```{r exponential2}
obj2 <- s_exponential(surv = 0.8, t = 1)
obj2

obj3 <- s_exponential(fail = 0.2, t = 1)
obj3
```

The following code shows how to make 100 simulations of 1000 subjects
with an object of the SURVIVAL class. The red line is the value from the
distribution.

```{r exponential3,  fig.height=4, fig.width=7, fig.align='center'}
obj4 <- s_exponential(surv = 0.25, t = 10)
ggplot_survival_random(obj4, timeto=10, subjects=1000, nsim=100, alpha = 0.1)
```

## Weibull distribution

The canonical parameters of the Weibull distribution are `scale` and
`shape`. The `scale` carry on the information about the time units. The
`scale` parameter can be derived from the proportion surviving or
failing at a given time but the `shape` needs to be provided by the
user. Both `scale` and `shape` needs to be numbers bigger than 0. A
value of `shape` equal to 1 is similar to an exponential distribution
with `lambda` parameter equal to the scale. If the `shape` is bigger
than 1 the hazard is increasing which means more events at the end of
follow up, and if between 0 and 1 is decreasing which translate to more
events at the beginning of the time at risk.

The following code shows the effect of the shape parameter on
distributions with the same scale.

```{r weibull, fig.height=4, fig.width=7, fig.align='center'}
wobj1 <- s_weibull(scale = 3, shape = 0.5)
wobj2 <- s_weibull(scale = 3, shape = 1)
wobj3 <- s_weibull(scale = 3, shape = 1.5)

par(mfrow=c(2,3))
plot(
  wobj1$sfx,
  from = 0,
  to = 1,
  main = "Weibull with shape 0.5",
  xlab = "Time",
  ylab = "Proportion without events",
  ylim = c(0,1))
plot(
  wobj2$sfx,
  from = 0,
  to = 1,
  main = "Weibull with shape 1",
  xlab = "Time",
  ylab = "Proportion without events",
  ylim = c(0,1))
plot(
  wobj3$sfx,
  from = 0,
  to = 1,
  main = "Weibull with shape 1.5",
  xlab = "Time",
  ylab = "Proportion without events",
  ylim = c(0,1))
plot(
  wobj1$hfx,
  from = 0,
  to = 1,
  xlab = "Time",
  ylab = "hazard")
plot(
  wobj2$hfx,
  from = 0,
  to = 1,
  xlab = "Time",
  ylab = "hazard")
plot(
  wobj3$hfx,
  from = 0,
  to = 1,
  xlab = "Time",
  ylab = "hazard")
par(mfrow=c(1,1))
```

## Gompertz distribution

The Gompertz distribution have two canonical parameters, the `scale` and
the `shape`. The `scale` needs to be a number higher than zero, and
represents the hazard at time 0. The `shape` can be any real number.
Negative `shape` produce a decreasing hazard. Positive `shape` produces
a increasing hazard. If the `shape` is zero, the distribution is reduced
to an exponential distribution, but this is not implemented in this
package. Instead an error is produced.

Similarly to the other distributions, the `scale` can be derived from
the survival or failing proportion at a given time, but the `shape`
parameter needs to be provided.

The following graph shows the effect of the `scale` parameter on the
Gompertz distribution

```{r gomperz, fig.height=4, fig.width=7, fig.align='center'}

# define a function to generate and plot Gompertz distributions
plot_sfx_gompertz<- function(shape, scale = 3, timeto = 1){
  plot(
    s_gompertz(shape = shape, scale = scale)$sfx,
    from = 0,
    to = timeto,
    main = paste("Shape: ", shape),
    xlab = "Time",
    ylab = "Proportion without events",
    ylim = c(0,1)
    )
}

plot_hfx_gompertz<- function(shape, scale = 3, timeto = 1){
  plot(
    s_gompertz(shape = shape, scale = scale)$hfx,
    from = 0,
    to = timeto,
    xlab = "Time",
    ylab = "hazard",
    ylim = c(2,4)
    )
}

par(mfrow=c(2,4))
plot_sfx_gompertz(shape = -0.25)
plot_sfx_gompertz(shape = -0.10)
plot_sfx_gompertz(shape = 0.10)
plot_sfx_gompertz(shape = 0.25)
plot_hfx_gompertz(shape = -0.25)
plot_hfx_gompertz(shape = -0.10)
plot_hfx_gompertz(shape = 0.10)
plot_hfx_gompertz(shape = 0.25)
par(mfrow = c(1,1))
```

## Piecewise Exponential distribution

The Piecewise Exponential distribution is a very flexible distribution
where the hazard is treated as constant until a breaks occurs and the
value of a new hazard is used. The class implements two parameters the
`breaks` that defines the breaks points and the `hazards` that define
the hazard used until the break point time. The factory function will
provide a warning if the last break is not `Inf` as otherwise the
distribution is not completely defined.

The parameters `break = c(1,2,3,Inf), hazards = c(0.1,3,4,3)` implements
a distribution where the hazard is 0.1 until time 1, 3 from time 1 until
time 2, a hazard of 4 until time 3 and from that point a hazard of 3
again.

The distribution can be also defined with the proportion surviving or
failing, `breaks` and `segments`. In this case the `segments` are scaled
to create hazards that results in a specified proportion surviving or
failing at the last not Inf break point. For example the parameters
`surv = 0.2, breaks = c(1,2,3,Inf),  segments = c(1, 2, 3, 1)` will
scale the segments to hazards in way that at time = 3 the surviving
proportion is 0.2. See the following example

```{r piecewise, fig.height=6, fig.width=7, fig.align='center'}
pobj <- s_piecewise(surv = 0.2, breaks = c(1,2,3,Inf), segments = c(1,2,3,1))
pobj
pobj$sfx(3)
plot_survival(pobj, timeto = 3)
```

## Comparison of SURVIVAL objects

The function `compare_survival()` can produce a graphic comparison of two 
SURVIVAL objects. The objects no need to be from the same distribution family.

```{r compare, fig.height=6, fig.width=7, fig.align='center'}

cobj1<- s_exponential(lambda = 3)
cobj2<- s_gompertz(scale = 3, shape = 0.4)
compare_survival(cobj1, cobj2, timeto = 2)
```

## References
