---
title: "Simulation of survival times"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation of survival times}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(survobj)
library(survival)
```

## Introduction

Following @bender2003 and @leemis1987, simulation of survival times is possible
if there is a function that inverts the cumulative hazard ($H^{-1}$). Random survival times for a baseline distribution can be generated from a uniform distribution between 0-1 $U$ as:
$$ T = H^{-1}(-log(U)) $$
For a survival distribution object, this can be accomplished with the function `rsurv(s_object, n)` which will generate `n` number of random draws from the distribution `s_object`. All objects of the s_distribution family implement a function that inverts the survival time with the function `invCum_Hfx()`

The function `ggplot_survival_random()` helps to graph Kaplan-Meier graphs and cumulative hazard of simulated times from the distribution

```{r, fig.height=4, fig.width=7, fig.align='center'}
s_obj <- s_exponential(fail = 0.4, t = 2)
ggplot_survival_random(s_obj, timeto =2, subjects = 1000, nsim= 10, alpha = 0.3)
```

## Generation of Proportional Hazard times 

Survival times with hazard proportional to the baseline hazard can be simulated 
$$ T = H^{-1}\left(\frac{-log(U)}{HR}\right) $$ where $HR$ is a hazard ratio.

The function `rsurvhr(s_object, hr)` can generate random numbers with hazards proportional to the baseline hazard. The function produces as many numbers as the length of the hr vector.
for example:

```{r, fig.height=4, fig.width=7, fig.align='center'}
s_obj <- s_exponential(fail = 0.4, t = 2)
group <- c(rep(0,500), rep(1,500))
hr_vector <- c(rep(1,500),rep(2,500))
times <- rsurvhr(s_obj, hr_vector)
plot(survfit(Surv(times)~group), xlim=c(0,5))
```
The function `ggplot_survival_hr()` can plot simulated data under proportional hazard assumption.

```{r, fig.height=4, fig.width=7, fig.align='center'}
s_obj <- s_exponential(fail = 0.4, t = 2)
ggplot_survival_hr(s_obj, hr = 2, nsim = 10, subjects = 1000, timeto = 5)
```


## Generation of Acceleration Failure Times 

Survival times with accelerated failure time to the baseline hazard can be simulated
$$ T = \frac{H^{-1}(-log(U))}{AFT}$$ where $AFT$ is an acceleration factor, meaning for example an AFT of 2 has events two times quicker than the baseline

The function `rsurvaft(s_object, aft)` can generate random numbers accelerated by an AFT factor. The function produces as many numbers as the length of the aft vector.
for example:

```{r, fig.height=4, fig.width=7, fig.align='center'}
s_obj <- s_lognormal(scale = 2, shape = 0.5)
ggplot_survival_aft(s_obj, aft = 2, nsim = 10, subjects = 1000, timeto = 5)
```

In this example, the scale parameter of the Log-Normal distribution represents the
median time, and in this simulation an acceleration factor of 2 moves the median time from 2 to 1

## Generation of Extended Hazards times

The Proportional Hazards and the Accelerated Failure Time effects can be
combined into a single model. Following @chenjewell2001, the Extended
Hazards model defines the hazard as
$$ h^*(t) = HR \cdot AFT \cdot h_0(AFT \cdot t) $$
which gives the cumulative hazard $H^*(t) = HR \cdot H_0(AFT \cdot t)$ and
survival time
$$ T = \frac{H_0^{-1}\left(\dfrac{-log(U)}{HR}\right)}{AFT} $$
This model nests both models described above as special cases: setting
$AFT = 1$ recovers the Proportional Hazards model, and setting $HR = 1$
recovers the Accelerated Failure Time model.

Note that this differs from the Accelerated Hazards model of @chen2000,
$h^*(t) = h_0(\theta t)$, which rescales the baseline hazard in time without
the additional Jacobian factor. That model is obtained from the Extended
Hazards model above as the special case $HR = 1/AFT$ (with $\theta = AFT$).

The function `rsurveh(s_object, aft, hr)` generates random numbers under the
Extended Hazards model. The function produces as many numbers as the length
of the `aft`/`hr` vectors, which must be of the same length. For example:

```{r, fig.height=4, fig.width=7, fig.align='center'}
s_obj <- s_lognormal(scale = 2, shape = 0.5)
ggplot_survival_eh(s_obj, aft = 2, hr = 0.5, nsim = 10, subjects = 1000, timeto = 5)
```

## References