---
title: "Simulating trials with survival endpoints"
author: John Aponte
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating trials with survival endpoints}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

We can use objects of the class SURVIVAL to simulate surviving times in
clinical trials. We present one example in which we want to estimate the
empirical power to detect significantly a vaccine efficacy.

The empirical power is define as the percentage of times the p-value is
equal or below 0.05.

```{r setup}
library(survobj)
library(ggplot2)
library(survival)
```

## Empirical power for superiority

Assumptions:

-   We made 1000 simulations

-   There are 300 participants in each group, one group is control and
    the other is vaccinated

-   The vaccine efficacy is 40% (i.e the hazard ratio is 1-40/100 = 0.6)

-   The control group follows an exponential distribution with 40% of
    subjects having an event at time 12 months (365.25 days)

-   There is equal probability in both groups to lost to follow-up and
    do not finalize the expected follow-up time. It is expected that 5%
    of the subjects are lost to follow-up

-   The simulated data is analized using Cox regression.

-   We estimate the empirical power as the percentage of the simulations
    where the p-value of the coefficient for the group is 0.05 or lower.
    We present the empirical power and the distribution of estimated
    vaccine efficacy

```{r simulation 1, fig.align='center', fig.width= 7, fig.height=5}

# Number of simulations
nsim = 1000

# Participants in each group
nsubjects = 300

# Vaccine efficacy
ve = 40

# Hazard ratio
hr = 1-ve/100

# Follow-up time
ftime <- 12

# Fail events in controls 
fail_control = 0.4

# Lost to follow up
fail_lof = 0.05

# Define Object with exponential distribution for events in controls
s_events <- s_exponential(fail = fail_control, t = ftime)

# Define Object with exponential distribution for lost to follow-up
s_lof <- s_exponential(fail = fail_lof, t = ftime)

## Simulation

# Define the group for the subjects
group = c(rep(0, nsubjects), rep(1, nsubjects))
    
# Define the hazard ratio according to the group
hr_vector <- ifelse(group ==0,1,hr)

# Loop    
sim <- lapply(
  1:nsim,
  function(x){
    # Simulate survival times for event
    sim_time_event <- s_events$rsurvhr(hr_vector)
    
    # Simulate survival times for lost to follow-up
    sim_time_lof <- s_lof$rsurv(length(group))
    
    # Censor events at end of follow-up
    cevent <- censor_event(censor_time = ftime, time = sim_time_event, event = 1)
    ctime <- censor_time(censor_time = ftime, time = sim_time_event)
    
    # Censor events due to lost to follow-up
    event <- censor_event(censor_time = sim_time_lof, time = ctime, event = cevent)
    timeto <- censor_time(censor_time = sim_time_lof, time = ctime)
    
    # Analyze the data using cox regression
    reg <- summary(coxph(Surv(timeto, event)~ group))
    
    # Collect the information
    pval = reg$coefficients["group","Pr(>|z|)"]
    ve = (1- exp(reg$coefficients["group","coef"]))*100
    
    # return values
    return(data.frame(simid = x, pval,ve))
  }
)

# Join all the simulations in a single data frame
sim_df <- do.call(rbind, sim)

# Analyze the simulation
empirical_power = mean(sim_df$pval <= 0.05)
empirical_power

# Distribution of the estimated VEs
plot(density(sim_df$ve, adjust = 2), main = "Simulated VE")
```
